---
title: "Hands-on Exercise 1"
author: "Nguyen Trong Nhan"
format:
  html:
    theme: lux
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    css: styles.css
---

# 1.1 Getting Started

## 1.1.1 Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# 1.2 Introducing ggplot

## 1.2.1 R Graphics VS ggplot

First, let us compare how R Graphics, the core graphical functions of Base R and ggplot plot a simple histogram.

::: panel-tabset
## R Graphics

```{r}
hist(exam_data$MATHS)

```

## ggplot2

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 10,
                 boundary = 100,
                 colour = "black",
                 fill = "grey") +
  ggtitle("Distribution of Maths scores")
```
:::

As you can see that the code chunk is relatively simple if R Graphics is used.

# 1.3 Grammar of Graphics

Grammar of Graphics is a general scheme for data visualization which breaks up graphs into semantic components such as scales and layers.

In the nutshell, Grammar of Graphics defines the rules of structuring mathematical and aesthetic elements into a meaningful graph.

There are two principles in Grammar of Graphics, they are:

-   Graphics = distinct layers of grammatical elements

-   Meaningful plots through aesthetic mapping

A good grammar of graphics will allow us to gain insight into the composition of complicated graphics, and reveal unexpected connections between seemingly different graphics (Cox 1978). It also provides a strong foundation for understanding a diverse range of graphics. Furthermore, it may also help guide us on what a well-formed or correct graphic looks like, but there will still be many grammatically correct but nonsensical graphics.

## 1.3.1 A Layered Grammar of Graphics

A short description of each building block are as follows:

-   Data: The dataset being plotted.

-   Aesthetics take attributes of the data and use them to influence visual characteristics, such as position, colours, size, shape, or transparency.

-   Geometrics: The visual elements used for our data, such as point, bar or line.

-   Facets split the data into subsets to create multiple variations of the same graph (paneling, multiple plots).

-   Statistics, statiscal transformations that summarise data (e.g. mean, confidence intervals).

-   Coordinate systems define the plane on which data are mapped on the graphic. Themes modify all non-data components of a plot, such as main title, sub-title, y-aixs title, or legend background.

# 1.4 Essential Grammatical Elements in ggplot2: Data

```{r}
ggplot(data=exam_data) 
```

# 1.5 Essential Grammatical Elements in ggplot2: Aesthetics

The aesthetic mappings take attributes of the data and and use them to influence visual characteristics, such as position, colour, size, shape, or transparency. Each visual characteristic can thus encode an aspect of the data and be used to convey information.

All aesthetics of a plot are specified in the [aes()](https://ggplot2.tidyverse.org/reference/aes.html) function call.

Code chunk below adds the aesthetic element into the plot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

# 1.6 Essential Grammatical Elements in ggplot2: geom

## 1.6.1 Geometric Objects: geom_bar

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

## 1.6.2 Geometric Objects: geom_dotplot

In a dot plot, the width of a dot corresponds to the bin width (or maximum width, depending on the binning algorithm), and dots are stacked, with each dot representing one observation.

In the code chunk below, [geom_dotplot()](https://ggplot2.tidyverse.org/reference/geom_dotplot.html) of ggplot2 is used to plot a dot plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

The y scale is not very useful, in fact it is very misleading. The ***scale_y_continuous()*** is used to turn off the y-axis, and ***binwidth*** argument is used to change the binwidth to 2.5.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

## 1.6.3 Geometric Objects: geom_histogram

Plot a histogram for Maths scores

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram() 
```

## 1.6.4 Geometric Objects: Modify by changing geom()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

-   *bins* argument is used to change the number of bins to 20,

-   *fill* argument is used to shade the histogram with light blue color, and

-   *color* argument is used to change the outline colour of the bars in black

## 1.6.5 Geometric Objects: Modify by changing aes()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

Changes the interior colour of the histogram (i.e. fill) by using sub-group of aesthetic().

## 1.6.6 Geometric Objects: geom_density

[geom_density()](https://ggplot2.tidyverse.org/reference/geom_density.html) computes and plots [kernel density estimate](https://en.wikipedia.org/wiki/Kernel_density_estimation), which is a smoothed version of the histogram.

It is a useful alternative to the histogram for continuous data that comes from an underlying smooth distribution.

The code below plots the distribution of Maths scores in a kernel density estimate plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

The code chunk below plots two kernel density lines by using ***colour*** or ***fill*** arguments of ***aes()***

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

## 1.6.7 Geometric Objects: geom_boxplot

[geom_boxplot()](https://ggplot2.tidyverse.org/reference/geom_boxplot.html) displays continuous value list. It visualises five summary statistics (the median, two hinges and two whiskers), and all “outlying” points individually.

The code chunk below plots boxplots by using [geom_boxplot().](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

[**Notches**](https://sites.google.com/site/davidsstatistics/home/notched-box-plots) are used in box plots to help visually assess whether the medians of distributions differ. If the notches do not overlap, this is evidence that the medians are different.

The code chunk below plots the distribution of Maths scores by gender in notched plot instead of boxplot.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

## 1.6.8 Geometric Objects: geom_violin

[`geom_violin`](https://ggplot2.tidyverse.org/reference/geom_violin.html) is designed for creating violin plot. Violin plots are a way of comparing multiple data distributions. With ordinary density curves, it is difficult to compare more than just a few distributions because the lines visually interfere with each other. With a violin plot, it’s easier to compare several distributions since they’re placed side by side.

The code below plot the distribution of Maths score by gender in violin plot.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

## 1.6.9 Geometric Objects: geom_point

[`geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html) is especially useful for creating scatterplot.

The code chunk below plots a scatterplot showing the Maths and English grades of pupils by using `geom_point()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

## 1.6.10 Geometric Objects: Combined

The code chunk below plots the data points on the boxplots by using both `geom_boxplot()` and `geom_point()`.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

# 1.7 Essential Grammatical Elements in ggplot2: stat

The [Statistics functions](https://ggplot2.tidyverse.org/reference/index.html#stats) statistically transform data, usually as some form of summary. For example:

-   frequency of values of a variable (bar graph)

    -   a mean

    -   a confidence limit

-   There are two ways to use these functions:

    -   add a `stat_()` function and override the default geom, or

    -   add a `geom_()` function and override the default stat.

## 1.7.1 Working with `stat()`

The boxplots below are incomplete because the positions of the means were not shown.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

## 1.7.2 Working with stat - the *stat_summary()* method

The code chunk below adds mean values by using [`stat_summary()`](https://ggplot2.tidyverse.org/reference/stat_summary.html) function and overriding the default geom.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```

## 1.7.3 Working with stat - the `geom()` method

The code chunk below adding mean values by using `geom_()` function and overriding the default stat.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```

## 1.7.4 Adding a best fit curve on a scatterplot?

The scatterplot below shows the relationship of Maths and English grades of pupils. The interpretability of this graph can be improved by adding a best fit curve.

In the code chunk below, [`geom_smooth()`](https://ggplot2.tidyverse.org/reference/geom_smooth.html) is used to plot a best fit curve on the scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

The default method used is *loess*. The default smoothing method can be overridden as shown below.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

# 1.8 Essential Grammatical Elements in ggplot2: Facets

Facetting generates small multiples (sometimes also called trellis plot), each displaying a different subset of the data. They are an alternative to aesthetics for displaying additional discrete variables. ggplot2 supports two types of factes, namely: [`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) and [`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html).

## 1.8.1 Working with `facet_wrap()`

[`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) wraps a 1d sequence of panels into 2d. This is generally a better use of screen space than facet_grid because most displays are roughly rectangular.

The code chunk below plots a trellis plot using `facet-wrap()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

## 1.8.2 Working with `facet_grid()`

[`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) forms a matrix of panels defined by row and column facetting variables. It is most useful when you have two discrete variables, and all combinations of the variables exist in the data.

The code chunk below plots a trellis plot using `facet_grid()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

# 1.9 Essential Grammatical Elements in ggplot2: Coordinates

## 1.9.1 Working with Coordinate

By the default, the bar chart of ggplot2 is in vertical form.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

The code chunk below flips the horizontal bar chart into vertical bar chart by using `coord_flip()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

## 1.9.2 Changing the y- and x-axis range

The scatterplot on the right is slightly misleading because the y-aixs and x-axis range are not equal.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

The code chunk below fixed both the y-axis and x-axis range from 0-100.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

# 1.10 Essential Grammatical Elements in ggplot2: themes

Themes control elements of the graph not related to the data. For example:

-   background colour

-   size of fonts

-   gridlines

-   colour of labels

Built-in themes include: - `theme_gray()` (default) - `theme_bw()` - `theme_classic()`

A list of theme can be found at this [link](https://ggplot2.tidyverse.org/reference/ggtheme.html). Each theme element can be conceived of as either a line (e.g. x-axis), a rectangle (e.g. graph background), or text (e.g. axis title).

## 1.10.1 Working with theme

The code chunk below plot a horizontal bar chart using `theme_gray()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

A horizontal bar chart plotted using `theme_classic()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

A horizontal bar chart plotted using `theme_minimal()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

# **1.11 References**

-   Hadley Wickham (2023) [**ggplot2: Elegant Graphics for Data Analysis**](https://ggplot2-book.org/). Online 3rd edition.

-   Winston Chang (2013) [**R Graphics Cookbook 2nd edition**](https://r-graphics.org/). Online version.

-   Healy, Kieran (2019) [**Data Visualization: A practical introduction**](https://socviz.co/). Online version

-   [Learning ggplot2 on Paper – Components](https://henrywang.nl/learning-ggplot2-on-paper-components/)

-   [Learning ggplot2 on Paper – Layer](https://henrywang.nl/learning-ggplot2-on-paper-layer/)

-   [Learning ggplot2 on Paper – Scale](https://henrywang.nl/tag/learning-ggplot2-on-paper/)
